---
title: "Data 606-Lab 1: Intro to R"
author: "Sean Connin"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### Exercise 1

```{r view-girls-counts}
arbuthnot$girls

ggplot(data=arbuthnot, aes(x=year, y=girls))+geom_point()



```


### Exercise 2

The number of girls baptized each year declined during the period 1640 and 1660 (to < 3000) before increasing again to levels >7000 after 1680. 

This pattern is apparent in the annual counts for boy baptisms as well. 

Annual boys:girls counts vary (without obvious temporal pattern) between 1.04 and 1.16 over the same time-span. More boys are baptized each year than girls. 
  

```{r trend-girls}
arbuthnot$boys+arbuthnot$girls

# create new column to sum boys + girls count by year

arbuthnot <- arbuthnot%>%mutate(total=girls+boys)

# create scatterplot of total baptism counts by year

ggplot(data=arbuthnot, aes(x=year, y=total))+geom_line()

#create scatterplot of baptism counts (boys) by year

ggplot(data=arbuthnot, aes(x=year, y=boys))+geom_line()

#create scatterplot of baptism counts (girls) count by year

ggplot(data=arbuthnot, aes(x=year, y=girls))+geom_line()

# create new tbl column with ratio boys:girls

arbuthnot<- arbuthnot%>%mutate(boy_to_girl_ratio=boys/girls)

#create scatterplot for the ratio of boy:girl baptism counts by year

ggplot(data=arbuthnot, aes(x=year, y=boy_to_girl_ratio))+geom_point()

# create new tbl column for the ratio of boy baptisms (to total) each year

arbuthnot <- arbuthnot%>%mutate(boy_ratio=boys/total)


```


### Exercise 3

It is difficult to discern a obvious trend(s) in the proportion of boys born over time. This owes, in part, to high inter-annual variation. 

Note: I don't believe we have established that baptism counts accurately track male/female births year-to-year.  

```{r plot-prop-boys-arbuthnot}


#create scatterplot for the ratio of boy baptism counts (to total) each year

ggplot(data=arbuthnot, aes(x=year, y=boy_ratio))+geom_point()

# evaluate number of births of boys relative to girls each year (see also: Exercise 2)

arbuthnot<- arbuthnot%>%mutate(more_boys=boys>girls)

# find the min and max count for boys

arbuthnot%>%summarize(min=min(boys), max=max(boys))
```


### Exercise 4

Included Years: 1940-202

DF dimensions: 63 rows, 3 cols

Column names: year, boys, girls

```{r dim-present}

#preview data set

present

# return years column

present$year

# return dataframe dimensions

dim(present)

# return column names

names(present)


```

### Exercise 5

Annual birth counts in the 'Present' data set are several orders of magnitude greater than annual baptism counts included in the Arbuthnot data set. 

Each data set covers a similar period of time, ~60 yrs. 

In the 'Present' data set, total birth counts increased by 1661327 between 1940 & 2002. Interim trends in birth counts were equivalent for boys and girls. 

There was notable, but temporary, decline in birth rates that culminated in the mid-70s. I speculate that this may related to economic stresses related to the oil crisis, etc. 

```{r count-compare}

# return max and min birth counts for boys and girls over period of record

present%>%pivot_longer(cols=c(boys, girls), names_to = 'gender', values_to ='value')%>%group_by(gender)%>%summarize(min=min(value), max=max(value))

# plot birth count of boys vs girls by year

plot_gender <- present%>%pivot_longer(cols=c(boys, girls), names_to = 'gender', values_to ='count')

ggplot(plot_gender, aes(x=year, y=count, color=gender)) +geom_line()


# create new column for total birth counts, plot counts by year, return difference in total count between 1940 & 2002

present <- present%>%mutate(total=girls+boys)

plot_total <- ggplot(data=present, aes(x=year, y=total))+geom_line()

present$total%>%last()-present$total%>%first()






```


### Exercise 6

Arbuthnot's observation that boys are born in greater proportion than girls in the U.S. does hold true in the 'Present' dataset.

The offset in boy and girl birth counts (relative to total) is similar in magnitude across data sets. 

However in the 'Present' data set, the proportion of male births (relative to total) decreases over time, while the opposite is true for females.

```{r plot-prop-boys-present}

# create new column with ratio of boys:total births by year, repeat  for girls, plot ratios for each gender.

present<- present%>%mutate(ratio_boys=boys/total)%>%mutate(ratio_girls=girls/total)

plot_ratio <- present%>%pivot_longer(cols=c(ratio_boys, ratio_girls), names_to = 'gender', values_to ='ratio_to_total')

ggplot(plot_ratio, aes(x=year, y=ratio_to_total, color=gender))+geom_line()
```


### Exercise 7

The greatest total number of births in the U.S. (for the period of record) occurred in 1961.

```{r find-max-total}

present %>%arrange(desc(total))

```

